### 2016-06-24 reasoning.tex
RG section made succinct.

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away the
environmental interference as a rely relation. A conspicuous
difference between a standard development of rely-guarantee and ours
is that, while the former reasons in terms of program states (variable
to value bindings), we reason in terms of executions as captured by
their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and pre ($P$) and post ($Q$) conditions are assertions over executions
(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on state are also written
similarly. For instance, consider the postcondition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is the invariant at a program point
inside \C{Wd1} described abstractly in \S\ref{sec:motivation}, and
reified as an execution-based assertion below:
\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{C}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E}(C) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E}(C) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}
One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
fact which cannot be deduced solely from the value of \C{C} (esp. in
presence of interference), but can be expressed as a proposition over
the execution trace.

@@

The standard rely guarantee judgment is the quintuple $\rg{P,R}{c}{G,Q}$,
which intuitively means the following: if a command $c$ is executed in
a state that satisfies its precondition $P$, provided that every
interference step during the execution is contained inside the rely
relation $R$, then the effect that each step of executing $c$ has on
the state is captured by $G$, and the final state of execution
satisfies the postcondition $Q$. The terms ``execution'' and ``step''
are defined with respect to an operational semantics, which in our
case, is parameterized on the trace invariant $\I$. Consequently, our
rely-guarantee judgment for commands is the sextuple\footnote{$\I$ is
written in the same group as $P$ to suggest that it is an assumption;
not an obligation.} $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs
from the quintuple in that it requires every step of the execution to
preserve the trace invariant $\I$ for the postcondition to be valid.
As usual, we write $\txnbox{t}_i$ to denote a term $t$ inside a
transaction $T_i$.  \txnimp expressions are side-effecting (they
generate \C{RD} effects), and admit interference during their
evaluation. Therefore, like the commands, expressions need to be
reasoned about explicitly. But, unlike commands, expressions evaluate
to values, and reasoning framework should also admit assertions on such
values. We therefore define a separate judgment for expressions - the
septuple $\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N}
\rightarrow \Prop$ is an assertion on the return value of $e$.

The rules that define rely-guarantee judgment are shown in
Fig.~\ref{fig:rg-rules}. Like the standard definitions of rely
guarantee, our definition also relies on the \emph{stability}
condition, which requires pre and post conditions to hold despite the
interference from concurrent threads (captured by $R$). The stability
of pre and post conditions can be predicated on the assumption that
the interference preserves the trace invariant $\I$. Formally:
\begin{smathpar}
\begin{array}{lcl}
  \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
  \I(\E) \conj P(\E) \conj R(\E,\E') \\
  &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
\end{array}
\end{smathpar}
However, the assumption that the interference preserves $\I$ (or,
equivalently, $\I$ withstands inteferference) needs to
be justified separately. We call this the stability requirement on
$\I$:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\I)& \Leftrightarrow & \forall \E, \E'.\, 
  \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
\end{array}
\end{smathpar}

The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre and post conditions to be stable relative to $\I$.
The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the precondition $P$, then the
resultant execution $\E'$ satisfies the postcondition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the
RG judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constitutent expressions $e_1$ and
$e_2$. Note that the pre-condition of $e_2$ ($Q'$) is also the
postcondition of $e_1$, reflecting the order of evaluation established
in Fig.~\ref{fig:txnimp}. The quantified premise requires any value
resulting from evaluating $e_1 \pm e_2$ to satisfy the assertion $C$,
provided that $e_1$ and $e_2$ always evaluate to values that satisfy
$C_1$ and $C_2$, respectively.

RG judgment for the assignment $X:=e$ makes use of the corresponding
judgment for the RHS expression $e$. The quantified premise asserts
that evaluating the assigment after evaluating $e$ to a value $v$ and
an execution $\E$, should result in an execution $\E'$ that satisfies
the postcondition $Q$, while being related to $\E$ via the guarantee
$G$. RG judgment for the \C{txn} lexical block is similar.  It uses
the judgment for the transaction-bound command $c$ (i.e.,
$\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
before the commit, and verifies that committing the transaction under
$\E$ results in an execution $\E'$ that satisfies transaction's
postcondition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
The rules for sequential composition of transaction-bound comands
(\rulelabel{RG-Seq}) and parallel composition of top-level commands
(\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
corresponding rules in classical rely-guarantee.

The three \rulelabel{RG-Conseq} rules define ways to strengthen or
weaken relations and assertions associated with RG judgments of
transaction-bound expressions, commands and top-level commands,
respectively. In particular, they allow the precondition $P$ and the
rely relation $R$ to be strengthened\footnote{A stronger relation is a
smaller set than a weaker relation, hence a stronger $R$ admits lesser
interference than a weaker $R$}, and the postcondition $Q$ (also, $C$
in the case of expressions) and the guarantee relation $G$ to be
weakened. This kind of strengthening/weakening is also admitted in the
classical RG framework. The most notable aspect of
\rulelabel{RG-Conseq} rules however is that they allow the trace
invariant $\I$ to be strengthened. Considering that $\I$ captures
consistency and isolation properties, this means that a program proved
correct under weaker consistency and isolation levels is also correct
under stronger levels.  Parametricity over the trace invariant $\I$,
combined with the ability to strengthen $\I$ as needed, allows our
proof system to admit a highly flexible proof strategy to prove
programs correct on various stores. Instead of requiring programs to
be proved correct with respect to a pre-defined store (\eg~A CC store
like in~\cite{gotsmanpopl16}), our reasoning framework allows
programmers to \emph{define} a store of their choice (by defining $\I$
appropriately) and then prove programs correct with respect to this
store. Thus, programmers can prove their programs correct with respect
to an imaginary store that provides only the required consistency and
isolation guarantees. Soundness of strengthening $\I$ guarantees that
the program can now be executed safely on any real store (\eg~a
relational database, or a NoSQL store) that offers same or stronger
consistency and isolation guarantees as the imaginary store. The
following subsection proves the soundness of our proof system.

@@

Soundness of rely-guarantee judgments can now be stated
thus\footnote{Supplemntary contains a rigorous proof of the soundness
theorem, and all relevant lemmas.}:
\begin{theorem}[Soundness] 
Rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to their semantics
defined above. 
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a). every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b). the effect
that every small-step reduction of $c$ along the way has on the trace
is contained in $G$. Soundness of RG judgment for transaction-bound
commands ($\txnbox{c}_i$) is stated similarly.  For expressions,
soundness of the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also
proves that $e$ is always evaluated to a value that satisfies $C$.

>>>>

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away the
environmental interference as a rely relation. A conspicuous
difference between a standard development of rely-guarantee and ours
is that, while the former reasons in terms of program states (variable
to value bindings), we reason in terms of executions as captured by
their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and pre ($P$) and post ($Q$) conditions are assertions over executions
(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on state are also written
similarly. For instance, consider the postcondition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:
\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{C}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E}(C) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E}(C) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}
One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
fact which cannot be deduced solely from the value of \C{C} (esp. in
presence of interference), but can be expressed as a proposition over
the execution trace.

@@

The standard rely guarantee judgment is the quintuple
$\rg{P,R}{c}{G,Q}$, which intuitively means the following: if a
command $c$ is executed in a state that satisfies its precondition
$P$, provided that every interference step during the execution is
contained inside the rely relation $R$, then the effect that each step
of executing $c$ has on the state is captured by $G$, and the final
state of execution satisfies the postcondition $Q$. The terms
``execution'' and ``step'' are defined with respect to an operational
semantics, which, in our case, is parameterized on the trace invariant
$\I$. Consequently, our rely-guarantee judgment for commands is the
sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
quintuple in that it requires every step of the execution to preserve
the trace invariant $\I$ for the postcondition to be valid.  As usual,
if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
expressions are side-effecting (they generate \C{RD} effects), and
admit interference during their evaluation.  Therefore, like the
commands, expressions need to be reasoned about explicitly. But,
unlike commands, expressions evaluate to values, and the reasoning
framework should also admit assertions on such values.  We therefore
define a separate judgment for expressions - the septuple
$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
\Prop$ is an assertion on the return value of $e$.

The rules that define rely-guarantee judgment are shown in
Fig.~\ref{fig:rg-rules}. Like the standard definitions of rely
guarantee, our definition also relies on the \emph{stability}
condition, which requires pre and post conditions to hold despite the
interference from concurrent threads (captured by $R$). The stability
of pre and post conditions can be predicated on the assumption that
the interference preserves the trace invariant $\I$. Formally:
\begin{smathpar}
\begin{array}{lcl}
  \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
  \I(\E) \conj P(\E) \conj R(\E,\E') \\
  &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
\end{array}
\end{smathpar}
However, the assumption that the interference preserves $\I$ (or,
dually, $\I$ withstands inteferference) needs to be justified
separately. We call this the stability requirement on $\I$:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\I)& \Leftrightarrow & \forall \E, \E'.\, 
  \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
\end{array}
\end{smathpar}

The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre and post conditions to be stable relative to $\I$.
The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the precondition $P$, then the
resultant execution $\E'$ satisfies the postcondition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the
RG judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constitutent expressions $e_1$ and
$e_2$. The quantified premise requires any value resulting from
evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
$C_2$, respectively. The rules for sequential and parallel composition
of comands are more-or-less same as the corresponding rules in
classical rely-guarantee, hence elided.

% RG judgment for the assignment $X:=e$ makes use of the corresponding
% judgment for the RHS expression $e$. The quantified premise asserts
% that evaluating the assigment after evaluating $e$ to a value $v$ and
% an execution $\E$, should result in an execution $\E'$ that satisfies
% the postcondition $Q$, while being related to $\E$ via the guarantee
% $G$. RG judgment for the \C{txn} lexical block is similar.  It uses
% the judgment for the transaction-bound command $c$ (i.e.,
% $\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
% before the commit, and verifies that committing the transaction under
% $\E$ results in an execution $\E'$ that satisfies transaction's
% postcondition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
% The rules for sequential composition of transaction-bound comands
% (\rulelabel{RG-Seq}) and parallel composition of top-level commands
% (\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
% corresponding rules in classical rely-guarantee.

The \rulelabel{RG-Conseq} rule defines ways to strengthen or weaken
relations and assertions associated with the RG judgment of
transaction-bound expressions. Similar rules exist for
transaction-bound and top-level commands, but are not shown. In
particular, the rules allow the precondition $P$ and the rely relation
$R$ to be strengthened\footnote{A stronger relation is a smaller set
than a weaker relation, hence a stronger $R$ admits lesser
interference than a weaker $R$}, and the postcondition $Q$ (also, $C$
in the case of expressions) and the guarantee relation $G$ to be
weakened. This kind of strengthening/weakening is also admitted in the
classical RG framework. The most notable aspect of
\rulelabel{RG-Conseq} rules however is that they allow the trace
invariant $\I$ to be strengthened. Considering that $\I$ captures
isolation properties, this means that a program proved correct under
weaker isolation levels is also correct under stronger levels.
Parametricity over the trace invariant $\I$, combined with the ability
to strengthen $\I$ as needed, allows our proof system to admit a
highly flexible proof strategy to prove programs correct on various
stores. For example, programmers can \emph{define} isolation
guarantees of their choice (by defining $\I$ appropriately) and then
prove programs correct assuming the guarantees hold. Soundness of
strengthening $\I$ assures that the program can now be executed safely
on any real store that offers same or stronger isolation guarantees
than those assumed. The following subsection proves the soundness of
our proof system.

@@

Soundness of rely-guarantee judgments can now be stated
thus\footnote{Supplementary contains a rigorous proof of the soundness
theorem, and all relevant lemmas.}:
\begin{theorem}[Soundness] 
Rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to their semantics
defined above. 
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a). every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b). the effect
that every small-step reduction of $c$ along the way has on the trace
is contained in $G$. Soundness of RG judgment for transaction-bound
commands ($\txnbox{c}_i$) is stated similarly.  For expressions,
soundness of the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also
proves that $e$ is always evaluated to a value that satisfies $C$.
